### 2025-03-13 paper/sections/method.qmd

::: {#def-explainability}

## Model Explainability

Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.

We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e., consistent with the data) and actionable. Formally, we define these properties as follows:

1. (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
2. (Actionability) Permutations $\Delta$ are subject to some actionability constraints.

We consider counterfactuals as faithful to the extent that they are consistent with what the model has learned about the input data. Let $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denote the conditional posterior over inputs, then formally:

3. (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.

:::

The characterization of faithfulness and plausibility in @def-explainability is the same as in @altmeyer2024faithful, with adapted notation. Actionability constraints in @def-explainability vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual. For our purposes, we let $T$ vary and consider the counterfactual search to be converged as soon as the predicted probability for the target class has reached a pre-determined threshold $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.]

To train models with high explainability as defined in @def-explainability, we propose to leverage counterfactuals in the following objective:

@@

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability. This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization but we depart from their methodology. Instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:

@@

Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the counterfactuals get more plausible (@def-explainability) during training, these opposing effects gradually balance each other out [@lippe2024uvadlc].

@@

::: {#prp-mtblty}

## Protecting Immutable Features

Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ (i.e., *multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.

:::

It is worth highlighting that @prp-mtblty assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

## Example: influence of counterfactual training on explainability

To better convey the intuition underlying our proposed method, we illustrate different model outcomes in @exm-grad.

::: {#exm-grad}

## Prediction of Consumer Credit Default

@@

In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time. The generated counterfactuals are all valid, but not plausible: they do not comply with the distribution of the factual samples in the target class to the point where they are clearly distinguishable from the ground-truth data. In panel (b), we have trained our model with counterfactual training, once again without any mutability constraints. We observe that the counterfactuals are highly plausible, meeting the first objective of @def-explainability.

In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time. Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants). By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected the immutable feature as described in @sec-constraints. We observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous and do not disproportionately punish younger individuals. The counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to @def-explainability.

:::

![Illustration for @exm-grad of how CT improves model explainability.](/paper/figures/poc.svg){#fig-poc}

>>>>

\begin{definition}[Model Explainability]
\label{def-explainability}
Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.

We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. Formally, we define these properties as follows,

\begin{enumerate}
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
\end{enumerate}

where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
\end{definition}

The characterization of faithfulness and plausibility in Def. \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about input data. Actionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual. For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see supplementary appendix.]

<!-- @sec-app-ce -->

To train models with high explainability as defined in Def. \ref{def-explainability}, we propose to leverage counterfactuals in the following objective:

@@

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in Def. \ref{def-explainability}. This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization but we depart from their methodology. Instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:

@@

Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the counterfactuals get more plausible (Def. \ref{def-explainability}) during training, these opposing effects gradually balance each other out [@lippe2024uvadlc].

@@

\begin{proposition}[Protecting Immutable Features]
\label{prp-mtblty}
Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
\end{proposition}

It is worth highlighting that Prp.~\ref{prp-mtblty} assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

## Example (Prediction of Consumer Credit Default)

@@

In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time. The generated counterfactuals are all valid, but not plausible: they do not comply with the distribution of the factual samples in the target class to the point where they are clearly distinguishable from the ground-truth data. In panel (b), we have trained our model with counterfactual training, once again without any mutability constraints. We observe that the counterfactuals are highly plausible, meeting the first objective of Def. \ref{def-explainability}.

In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time. Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants). By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected the immutable feature as described in @sec-constraints. We observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous and do not disproportionately punish younger individuals. The counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

![Illustration of how CT improves model explainability.](/paper/figures/poc.svg){#fig-poc}

### 2025-03-13 paper/sections/other/constraints.qmd

In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature. We argue in @sec-constraints that this approach induces models to be less sensitive to immutable features and demonstrate this empirically in @sec-experiments. Below we derive the analytical results in @prp-mtblty.

@@

::: {.callout-warning}

\@Cynthia, \@Arie, I have tentatively phrased the above in terms of a theorem and proof. This is something I've so far shied away from because I feel a bit out of my depth when it comes to mathematical proofs. The above makes intuitive sense to me, but I don't know for sure if it's correct.

>>>>

In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature. We argue in @sec-constraints that this approach induces models to be less sensitive to immutable features and demonstrate this empirically in @sec-experiments. Below we derive the analytical results in Prp.~\ref{prp-mtblty}.

@@
